---
title: "Environmental footprint analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Environmental footprint analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## What is a footprint analysis?

A **footprint analysis** answers the question: *"How much of an
environmental pressure (land, water, emissions...) caused in country A
is ultimately driven by consumption in country B?"*

Modern supply chains are global. Spain might import soybeans from Brazil,
feed them to pigs, and export ham to France. A footprint analysis traces
these flows backwards through the entire supply chain, so that the
environmental cost of growing those soybeans is attributed to the French
consumer who ate the ham — not to the Brazilian farmer who grew the
beans.

This is done by building a **Multi-Regional Input-Output (MRIO) model**,
a mathematical framework that captures all inter-industry and
inter-country linkages in one system of equations.

## The three-step pipeline

The whep package implements this in three steps:

1. **`build_io_model()`** — Assemble the MRIO tables from supply-use
   data, bilateral trade, and commodity balance sheets.
2. **`compute_leontief_inverse()`** — Compute the Leontief inverse,
   which captures both direct and indirect supply chain linkages.
3. **`compute_footprint()`** — Multiply the Leontief inverse by
   environmental extension data to trace pressures through to final
   consumption.

The methodology follows the
[FABIO model](https://doi.org/10.1021/acs.est.9b03554) (Bruckner et
al., 2019), which builds physical MRIO tables for the global food
system.

## Concepts and terminology

Before diving into the code, here are the key concepts:

### Supply-use tables

A **supply table** records how much of each product each process
produces (e.g. wheat milling produces flour). A **use table** records
how much of each product each process consumes (e.g. bakeries consume
flour). Together they describe the technology of the economy: who
produces what, and who uses what as inputs.

### Bilateral trade

A **bilateral trade matrix** records, for each product, how many tonnes
flow from each exporting country to each importing country. This is what
connects domestic economies into a global system.

### Commodity balance sheets (CBS)

A **commodity balance sheet** is an accounting identity for each product
in each country:

$$
\text{Production} + \text{Imports} = \text{Exports}
  + \text{Food} + \text{Feed} + \text{Other uses}
  + \Delta\text{Stocks}
$$

The right-hand categories (food, feed, other uses) form the
**final demand** — the end of the supply chain where products are
consumed rather than being used as intermediate inputs.

### The Z matrix (inter-industry flows)

`Z` is the core of the MRIO model. It is a square matrix where each row
and column represents a *(country, product)* pair. Entry $Z_{ij}$
gives the physical flow from sector $i$ to sector $j$.

For example, if row 5 is *(Spain, wheat)* and column 12 is *(Spain,
flour)*, then $Z_{5,12}$ is the amount of Spanish wheat used by
Spanish flour mills.

The model constructs `Z` using the **industry technology assumption**:

$$
Z = U_{MR} \times T
$$

where $U_{MR}$ is the multi-regional use matrix (expanded with trade
shares) and $T$ is the row-normalized supply matrix (the transformation
matrix). This means: "the inputs each process uses are distributed
across origin countries proportionally to trade shares, and then
allocated to products proportionally to how each process distributes its
output."

### Trade shares

For each product, trade shares tell us *where each country gets its
supply from*. If Spain consumes 1000 tonnes of soybeans, and 600 come
from domestic production while 400 come from Brazil, then the trade
share from Brazil to Spain is 0.4 and the domestic share is 0.6.

These shares are computed from the bilateral trade matrices combined
with domestic production data from the CBS. They are used to expand the
single-country use tables into a multi-regional system where each
input can originate from any country.

### The Y matrix (final demand)

`Y` is the final demand matrix. Rows are *(country, product)* pairs
(same as `Z`), and columns are *(country, demand category)* pairs.
Entry $Y_{ij}$ tells us how much of product $i$ goes to final demand
category $j$.

Like `Z`, the final demand is expanded with trade shares so that, for
example, food consumption of wheat in France can partly originate from
French wheat and partly from imported wheat.

### The X vector (total output)

`X` is a vector where $X_i$ is the total output of sector $i$.
By construction:

$$
X = Z \cdot \mathbf{1} + Y \cdot \mathbf{1}
$$

meaning total output equals intermediate deliveries plus final demand.

### The Leontief inverse (L)

The **Leontief inverse** $L = (I - A)^{-1}$ is the key analytical
tool. The technical coefficients matrix $A$ is defined as
$A_{ij} = Z_{ij} / X_j$, representing how much input from sector $i$
is needed per unit of output from sector $j$.

To ensure the system is invertible even when supply-use data contain
inconsistencies (e.g. input flows exceed reported output), column sums
of $A$ are capped at 1. This is the standard FABIO convention.

The Leontief inverse captures the **total requirements** — both direct
and indirect — needed to deliver one unit of final demand. Entry
$L_{ij}$ tells us: *"To deliver one extra unit of product $j$ to final
consumers, how much total output does sector $i$ need to produce
(including all the intermediate steps)?"*

This is what makes footprint tracing possible. A consumer buying bread
in France triggers not just French bakeries, but also French mills,
French wheat farmers, Brazilian soybean exporters (if feed is
involved), and so on. The Leontief inverse encodes all of these ripple
effects in a single matrix.

### Environmental extensions

An **environmental extension** is any pressure you want to trace through
the supply chain. Common examples:

- **Land use** (hectares): How many hectares of cropland does each
  sector use?
- **Water use** (cubic meters): How much blue water is consumed?
- **Greenhouse gas emissions** (tonnes CO₂-eq): How much does each
  sector emit?
- **Nitrogen surplus** (tonnes N): How much excess nitrogen enters the
  environment?

You provide these as a vector $e$ with one value per *(country,
product)* sector.

### The footprint

The footprint matrix is computed as:

$$
FP = \hat{f} \cdot L \times Y
$$

where $\hat{f}$ is a diagonal matrix of **extension intensities**
$f_i = e_i / X_i$ (pressure per unit of output). Each entry $FP_{ij}$
tells us: *"How much environmental pressure occurring in sector $i$ is
driven by final demand category $j$?"*

The resulting tidy tibble decomposes this by origin country/product and
target country/product, giving you a full bilateral footprint map.

## Running the analysis

### Step 0: Prepare the input data

You need three datasets, all of which can be built with whep functions:

```{r}
library(whep)

# Supply-use tables (who produces and uses what)
supply_use <- build_supply_use()

# Bilateral trade matrices (who trades with whom)
bilateral_trade <- get_bilateral_trade()

# Commodity balance sheets (production, consumption, trade)
cbs <- get_wide_cbs()
```

### Step 1: Build the IO model

```{r}
# For all years in common (default)
io <- build_io_model(supply_use, bilateral_trade, cbs)

# Or filter to specific years (useful for large timeseries)
io <- build_io_model(
  supply_use, bilateral_trade, cbs, years = c(2010, 2015, 2020)
)
io
```

This returns a tibble with one row per year. Each row contains:

- `Z`: The inter-industry flow matrix.
- `Y`: The final demand matrix.
- `X`: The total output vector.
- `labels`: A tibble mapping matrix indices to `area_code` and
  `item_cbs_code`.

You can inspect a single year:

```{r}
# Extract matrices for one year
z_mat <- io$Z[[1]]
y_mat <- io$Y[[1]]
x_vec <- io$X[[1]]
labels <- io$labels[[1]]

# See dimensions: (n_countries * n_products) x (n_countries * n_products)
dim(z_mat)

# See label mapping
labels
```

### Step 2: Compute the Leontief inverse

For **small systems** (< 5000 sectors), you can pre-compute L:

```{r}
l_inv <- compute_leontief_inverse(z_mat, x_vec)
dim(l_inv)
```

This inverts the $(I - A)$ matrix. Column sums of $A$ are internally
capped at 1 to prevent singularity.

#### For large systems

For systems larger than ~5000 sectors (187 countries × 125+ products),
computing the explicit Leontief inverse requires too much memory
(e.g. 4.8 GiB for 25,000 sectors). Instead, use the **sparse path** in
Step 3 below, which solves $(I - A) x = Y$ without ever materialising
the dense matrix $L$.

### Step 3: Compute the footprint

Now provide your environmental extension vector. This must have one
value per sector, in the same order as the labels. For example, if you
have land use data:

```{r}
# Suppose you have a tibble with land use per country and product
land_use_data <- tibble::tibble(
  area_code = c(1L, 1L, 2L, 2L),
  item_cbs_code = c(2511L, 2514L, 2511L, 2514L),
  hectares = c(1e6, 5e5, 2e6, 3e5)
)

# Match it to the IO model's label order
extensions <- labels |>
  dplyr::left_join(
    land_use_data,
    by = c("area_code", "item_cbs_code")
  ) |>
  dplyr::mutate(
    hectares = tidyr::replace_na(hectares, 0)
  ) |>
  dplyr::pull(hectares)
```

#### Dense path (pre-computed L)

If you pre-computed L:

```{r}
footprint <- compute_footprint(
  l_inv, x_vec, y_mat, extensions, labels
)
footprint
```

#### Using Z directly (computes L internally)

Alternatively, pass the Z matrix and `compute_footprint()` will
derive the Leontief inverse for you:

```{r}
footprint <- compute_footprint(
  z_mat = z_mat,
  x_vec = x_vec,
  y_mat = y_mat,
  extensions = extensions,
  labels = labels
)
footprint
```

#### Per-item decomposition with fd_labels

Pass `fd_labels` from the IO model to decompose the footprint per
target item and demand category, following the FABIO approach:

```{r}
footprint <- compute_footprint(
  z_mat = z_mat,
  x_vec = x_vec,
  y_mat = y_mat,
  extensions = extensions,
  labels = labels,
  fd_labels = fd_labels
)
footprint
```

When `fd_labels` is provided, the result includes `target_item`
(the consumed product) and `target_fd` (the demand category like
`"food"` or `"feed"`).

The result is a tidy tibble with columns:

| Column | Meaning |
|--------|---------|
| `origin_area` | Country where the pressure physically occurs |
| `origin_item` | Product causing the pressure |
| `target_area` | Country whose consumption drives the pressure |
| `target_item` | Product being consumed |
| `value` | Footprint in extension units (e.g. hectares) |

### Analysing the results

Because the output is a tidy tibble, you can use standard dplyr
operations to answer questions:

**Which country's consumption drives the most land use?**

```{r}
footprint |>
  dplyr::summarise(
    total_land = sum(value), .by = target_area
  ) |>
  dplyr::arrange(dplyr::desc(total_land))
```

**How much of Spain's land use is driven by foreign consumption?**

```{r}
spain_code <- 203L

footprint |>
  dplyr::filter(origin_area == spain_code) |>
  dplyr::mutate(
    destination = ifelse(
      target_area == spain_code, "domestic", "exported"
    )
  ) |>
  dplyr::summarise(
    total_land = sum(value), .by = destination
  )
```

**What are the top bilateral flows?**

```{r}
footprint |>
  dplyr::summarise(
    total = sum(value),
    .by = c(origin_area, target_area)
  ) |>
  dplyr::arrange(dplyr::desc(total)) |>
  head(20)
```

### A conservation check

A useful sanity check: the total footprint should equal the total
extensions. All pressure must be attributed somewhere.

```{r}
# These two numbers should be equal
sum(footprint$value)
sum(extensions)
```

## Processing multiple years

Since `build_io_model()` returns one row per year, you can filter to
specific years at the model level, or loop over them:

#### Fast approach: filter years early

```{r}
# Only build the years you need
io <- build_io_model(
  supply_use, bilateral_trade, cbs, years = 2010:2020
)

all_footprints <- purrr::pmap_dfr(
  list(
    io$Z, io$X, io$Y, io$labels,
    io$fd_labels, io$year
  ),
  function(z, x, y, lab, fdl, yr) {
    ext <- rep(0, length(x))  # Replace with your extension data
    compute_footprint(
      z_mat = z, x_vec = x, y_mat = y,
      extensions = ext, labels = lab,
      fd_labels = fdl
    ) |>
      dplyr::mutate(year = yr)
  }
)
```

#### Dense path (if system is small)

If your system has < 5000 sectors, you can pre-compute L:

```{r}
all_footprints <- purrr::pmap_dfr(
  list(
    io$Z, io$X, io$Y, io$labels,
    io$fd_labels, io$year
  ),
  function(z, x, y, lab, fdl, yr) {
    l_inv <- compute_leontief_inverse(z, x)
    ext <- rep(0, length(x))  # Replace with your extension data
    compute_footprint(
      l_inv, x, y, ext, lab,
      fd_labels = fdl
    ) |>
      dplyr::mutate(year = yr)
  }
)
```

#### Parallel execution (optional)

Since each year is independent, you can parallelise using `furrr`:

```{r}
library(furrr)
future::plan(multisession, workers = 4)

all_footprints <- furrr::future_pmap_dfr(
  list(
    io$Z, io$X, io$Y, io$labels,
    io$fd_labels, io$year
  ),
  function(z, x, y, lab, fdl, yr) {
    ext <- rep(0, length(x))
    compute_footprint(
      z_mat = z, x_vec = x, y_mat = y,
      extensions = ext, labels = lab,
      fd_labels = fdl
    ) |>
      dplyr::mutate(year = yr)
  }
)
```

## Summary

| Step | Function | Input | Output |
|------|----------|-------|--------|
| 1 | `build_io_model()` | Supply-use, trade, CBS | Z, Y, X matrices per year |
| 2 | `compute_leontief_inverse()` | Z, X | Leontief inverse L |
| 3 | `compute_footprint()` | L, X, Y, extensions, labels | Tidy bilateral footprint |

## References

Bruckner, M., Wood, R., Moran, D., Kuschnig, N., Wieland, H.,
Maus, V., & Börner, J. (2019). FABIO — The Construction of the Food
and Agriculture Biomass Input-Output Model. *Environmental Science &
Technology*, 53(19), 11302–11312.
https://doi.org/10.1021/acs.est.8b03704
