---
title: "Footprint Sankey viewer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Footprint Sankey viewer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(whep)
library(tibble)
```

The Sankey viewer is a browser-native HTML visualisation for footprint
path tables. It is suitable for pkgdown because the interactive HTML is
embedded in the article output and does not require a separate server.

This page uses a compact example so the documentation site remains fast
to build. Full one-year viewers can be generated separately with
`inst/scripts/footprint_sankey_one_year.R`.

```{r sankey-demo}
sankey_paths <- tribble(
  ~origin_area, ~origin_item, ~product_area, ~product, ~target_area, ~target_fd, ~value,
  "Brazil", "Soybeans", "China", "Pigmeat (China)", "China", "food", 820,
  "Brazil", "Soybeans", "China", "Poultry Meat (China)", "China", "food", 260,
  "Brazil", "Soybeans", "China", "Bovine Meat (China)", "China", "food", 180,
  "Brazil", "Soybeans", "China", "Milk - Excluding Butter (China)", "China", "food", 115,
  "Brazil", "Soybeans", "Brazil", "Soybean Oil (Brazil)", "Brazil", "food", 210,
  "Brazil", "Soybeans", "Brazil", "Soybean Cake (Brazil)", "Brazil", "feed", 170,
  "Brazil", "Soybeans", "Spain", "Pigmeat (Spain)", "Spain", "food", 95,
  "Brazil", "Soybeans", "France", "Poultry Meat (France)", "France", "food", 70,
  "Brazil", "Soybeans", "Rest of world", "Agricultural products, nes (Rest of world)",
  "Rest of world", "food", 140
)

plot_footprint_sankey(
  sankey_paths,
  stages = c(
    "origin_area",
    "origin_item",
    "product_area",
    "product",
    "target_area",
    "target_fd"
  ),
  max_nodes = 8,
  stage_max_nodes = c(product = 12, target_fd = 3),
  min_share = 0,
  title = "Brazil Soybeans Footprint Sankey",
  subtitle = "Compact example of origin products traced through supplied products to final demand.",
  value_label = "ha",
  height = 560
)
```

The same viewer can be written as a standalone HTML file:

```{r standalone-example, eval = FALSE}
plot_footprint_sankey(
  sankey_paths,
  stages = c(
    "origin_area",
    "origin_item",
    "product_area",
    "product",
    "target_area",
    "target_fd"
  ),
  file = "footprint-sankey-example.html",
  value_label = "ha"
)
```

For a full model year, generate the footprint paths first and let the
script write a standalone viewer. The script embeds at most 20 individual
nodes per stage by default; set `WHEP_FOOTPRINT_SANKEY_EMBED_MAX_NODES=Inf`
to write the full path detail.

```{bash full-year-example, eval = FALSE}
Rscript inst/scripts/footprint_sankey_one_year.R 2000 footprint-sankey-2000.html
```
