---
title: "How areas resolve to polities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How areas resolve to polities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(whep)
```

Every area, country and region identifier in this package resolves to a **polity** — a
territorial-economic unit from the `whep-polities` database, coded `PREFIX-startyear-endyear`.
This vignette documents how that resolution works, which columns are safe to join on, and
where the honest gaps are.

## What the package embeds, and what it verifies

`whep-polities` publishes four files. This package embeds a copy of each and checks it has not
drifted:

| upstream file | embedded as | what is verified |
|---|---|---|
| `polities_database.gpkg` | `polities` | row-by-row on eight identity fields |
| `faostat_area_polity_map.csv` | `polity_area_crosswalk` | no area conflicts on overlapping years |
| `label_alias_map.csv` | `polity_label_aliases` | every alias targets a polity this package carries |
| `polities_manifest.json` | — | identity fields, dead codes, polygon gaps |

Those comparisons need the upstream repository, which is private, so they skip on CI. What runs
everywhere is the comparison **between the two embedded objects**: `polities` and
`polity_area_crosswalk` are built by separate steps from the same source, so disagreement on any
identity field means one was regenerated and the other was not.

## Resolving an area

```{r}
add_polity_code(data.frame(area_code = c(185L, 51L), year = c(2000L, 1960L)))[
  , c("area_code", "year", "polity_code")
]
```

Two conventions matter:

- **`end_year` is exclusive.** Area 185 in 2014 resolves to `RUS-2014-2025`, not `RUS-1991-2014`.
- **Years before 1961 match the polity active in 1961.** Not a convenience: WHEP's pre-1961
  series are *back-cast onto 1961 territory* rather than reported under their own year's borders,
  so the 1961-active polity is the correct target and matching by data year would attribute
  1961-border figures to an entity with different borders. On the real production build this
  covers 1,978,400 rows — 31% of the dataset — whose polity span excludes the row's own year.
  Pass `backcast_anchor = -Inf` to match strictly by data year instead.

A resolved row carries both a code and a numeric key, and they answer different questions:

- `reporting_polity_code` — the polity. Join this to `polities`.
- `polity_area_code` — the numeric area key the matrix workflows use. Most builds emit this
  **as** their `area_code`, so it, not the polity code, determines how rows aggregate.

That distinction is load-bearing. Eleven areas that FABIO folds into rest-of-world report data of
their own, and correcting only their `polity_code` left them still summed into area 999 by the
numeric key.

## Resolving a label

`resolve_polity_label()` needs **both** a source and a year. Aliases are source-scoped and
year-ranged, so omitting either silently returns `NA` rather than erring:

```{r}
resolve_polity_label(
  c("Ivory Coast", "Türkiye", "Eswatini"),
  source = "faostat",
  year = 2000L
)
```

```{r}
# Without a year, nothing resolves — the aliases are all year-ranged.
resolve_polity_label("Ivory Coast", source = "faostat")
```

## Three columns look like an area name; one is safe

`regions_full` carries `name`, `FAOSTAT_name` and `polity_name`. Only one matches the crosswalk:

```{r}
cw <- unique(as.data.frame(polity_area_crosswalk)[, c("area_code", "area_name")])
rf <- as.data.frame(regions_full)[, c("code", "name", "FAOSTAT_name", "polity_name")]
both <- merge(cw[!is.na(cw$area_code), ], rf, by.x = "area_code", by.y = "code")
vapply(
  c("name", "FAOSTAT_name", "polity_name"),
  function(col) sum(both$area_name != both[[col]], na.rm = TRUE),
  integer(1)
)
```

`FAOSTAT_name` is canonical. `name` carries short forms (Turkey, Czech Republic, Ivory Coast) and
`polity_name` carries the aggregate label for folded areas (Bermuda reads "Latin America Other").
A lot of code joins on `c("year", "area", ...)`, and using either non-canonical column loses rows
silently — an unmatched join yields `NA`, not an error. Two inputs had exactly that defect:
`gdp-population` lost 6.4% of its rows and the fodder bridge 13.3%.

## Where the gaps are

Stated rather than hidden, because each is a real limit on what the data can say:

- **Rest-of-world folding.** Areas with no data of their own fold into a `ROW-*` polity, which is
  what FABIO does. Areas that *do* report data are excluded from the fold, decided by the
  `observed_rows` column of the published alias map.
- **Pre-1961 back-casting** needs a LUH2 land series per area. The French overseas departments
  have none — LUH2 folds *départements* into France — so their earlier production is not
  back-cast rather than being estimated from an aggregate.
- **IPCC regions** come from the GLEAM hierarchy, which does not cover 59 of the areas carrying
  an ISO3. Those use the `Global` coefficient, which covers every category.
- **Reporting buckets** (`RLAM-1850-2013` and the other regional "Other" rows) end before the
  data does, so their most recent years resolve to nothing.
- **Most of the crosswalk is unreachable, by design.** Of its 599 area-polity pairs, 274 can
  never be returned by `add_polity_code()`: 273 belong to polities that ended at or before 1961,
  and the clamp means no earlier year is matched directly. The crosswalk deliberately carries
  whole polity chains, while FAOSTAT-sourced resolution can only deliver the 1961-onward part of
  each. A mapping existing in the crosswalk is therefore not evidence that any query returns it.

```{r}
# The one unreachable pair that does NOT end before 1961 — area 15 maps to two overlapping
# Belgium-Luxembourg rows and the resolver only ever picks the first.
add_polity_code(data.frame(area_code = 15L, year = c(1930L, 1970L)))$polity_code
```
