---
title: "How areas resolve to polities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How areas resolve to polities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(whep)
```

Every area, country and region identifier in this package resolves to a **polity** — a
territorial-economic unit from the `whep-polities` database, coded `PREFIX-startyear-endyear`.
This vignette documents how that resolution works, which columns are safe to join on, and
where the honest gaps are.

## What the package embeds, and what it verifies

`whep-polities` publishes four files. This package embeds a copy of each and checks it has not
drifted:

| upstream file | embedded as | what is verified |
|---|---|---|
| `polities_database.gpkg` | `polities` | row-by-row on eight identity fields |
| `faostat_area_polity_map.csv` | `polity_area_crosswalk` | no area conflicts on overlapping years |
| `label_alias_map.csv` | `polity_label_aliases` | every alias targets a polity this package carries |
| `polities_manifest.json` | — | identity fields, dead codes, polygon gaps |

Those comparisons need the upstream repository, which is private, so they skip on CI. What runs
everywhere is the comparison **between the two embedded objects**: `polities` and
`polity_area_crosswalk` are built by separate steps from the same source, so disagreement on any
identity field means one was regenerated and the other was not.

## Resolving an area

```{r}
add_polity_code(data.frame(area_code = c(185L, 51L), year = c(2000L, 1960L)))[
  , c("area_code", "year", "polity_code")
]
```

Two conventions matter:

- **`end_year` is exclusive.** Area 185 in 2014 resolves to `RUS-2014-2025`, not `RUS-1991-2014`.
- **Years before 1961 match the polity active in 1961.** Not a convenience: WHEP's pre-1961
  series are *back-cast onto 1961 territory* rather than reported under their own year's borders,
  so the 1961-active polity is the correct target and matching by data year would attribute
  1961-border figures to an entity with different borders. On the real production build this
  covers 1,978,400 rows — 31% of the dataset — whose polity span excludes the row's own year.
  Pass `backcast_anchor = -Inf` to match strictly by data year instead.

A resolved row carries both a code and a numeric key, and they answer different questions:

- `reporting_polity_code` — the polity. Join this to `polities`.
- `polity_area_code` — the numeric area key the matrix workflows use. Most builds emit this
  **as** their `area_code`, so it, not the polity code, determines how rows aggregate.

That distinction is load-bearing. Eleven areas that FABIO folds into rest-of-world report data of
their own, and correcting only their `polity_code` left them still summed into area 999 by the
numeric key.

## Resolving a label

`resolve_polity_label()` needs **both** a source and a year. Aliases are source-scoped and
year-ranged, so omitting either silently returns `NA` rather than erring:

```{r}
resolve_polity_label(
  c("Ivory Coast", "Türkiye", "Eswatini"),
  source = "faostat",
  year = 2000L
)
```

```{r}
# Without a year, nothing resolves — the aliases are all year-ranged.
resolve_polity_label("Ivory Coast", source = "faostat")
```

## Three columns look like an area name; one is safe

`regions_full` carries `name`, `FAOSTAT_name` and `polity_name`. Only one matches the crosswalk:

```{r}
cw <- unique(as.data.frame(polity_area_crosswalk)[, c("area_code", "area_name")])
rf <- as.data.frame(regions_full)[, c("code", "name", "FAOSTAT_name", "polity_name")]
both <- merge(cw[!is.na(cw$area_code), ], rf, by.x = "area_code", by.y = "code")
vapply(
  c("name", "FAOSTAT_name", "polity_name"),
  function(col) sum(both$area_name != both[[col]], na.rm = TRUE),
  integer(1)
)
```

`FAOSTAT_name` is canonical. `name` carries short forms (Turkey, Czech Republic, Ivory Coast) and
`polity_name` carries the aggregate label for folded areas (Bermuda reads "Latin America Other").
A lot of code joins on `c("year", "area", ...)`, and using either non-canonical column loses rows
silently — an unmatched join yields `NA`, not an error. Two inputs had exactly that defect:
`gdp-population` lost 6.4% of its rows and the fodder bridge 13.3%.

## Datasets that are not keyed by a FAOSTAT area

Everything above is about `area_code`. But a census of the package's data — the lazy-loaded
datasets *and* the internal ones, since an `ls()` over the namespace alone misses lazy data and
would report a single table — finds 21 area-keyed tables with no polity column. They split cleanly
in two, and the split is the whole point: one half should carry polities and now does, the other
half should not.

**Keyed by a territory.** Each resolves through the published alias map, under a source name, with
year-scoped aliases maintained upstream rather than name-matched here:

```{r}
resolve_share <- function(labels, source, year) {
  round(100 * mean(!is.na(resolve_polity_label(labels, source = source, year = year))), 1)
}
lass <- unique(lassaletta_grassland_share[, c("Country", "year")])
by_year <- tapply(
  seq_len(nrow(lass)),
  lass$year,
  function(i) {
    !is.na(resolve_polity_label(
      as.character(lass$Country[i]),
      source = "lassaletta-grassland-share",
      year = lass$year[i][[1]]
    ))
  }
)
c(
  lassaletta_rows = nrow(lass),
  lassaletta_pct = round(100 * mean(unlist(by_year)), 1),
  crops_manure_pct = resolve_share(
    unique(stats::na.omit(crops_manure_n$ISO)),
    "crops-manure-n",
    2000L
  ),
  mueller_pct = resolve_share(
    unique(stats::na.omit(mueller_synthetic_n$iso3c)),
    "mueller-synthetic-n",
    2000L
  )
)
```

Mueller and `crops_manure_n` are complete, and Lassaletta is complete in the only sense available
to it: every one of its remaining unresolved country-years is a **source anachronism** — a label
carried outside the period its entity existed. `South Sudan` is listed from 1961 though it did not
exist until 2011; `FSU` and `Yugoslav SFR` run past their dissolutions; `Czechoslovakia` reaches
2009 though it dissolved in 1993; `Belgium-Luxemburg` continues past the 1999 split. Each resolves
correctly for the years its entity *did* exist, so resolving the tails would be wrong rather than
better.

`expand_trade_sources()` is the same shape, and complete: its rows are already one year each, so
the reporter and the year settle every row.

```{r}
src <- utils::read.csv(
  system.file("extdata", "trade_sources.csv", package = "whep"),
  check.names = FALSE
)
out <- expand_trade_sources(src)
c(
  rows = nrow(out),
  with_polity = sum(!is.na(out$reporting_polity_code)),
  distinct_polities = length(unique(out$reporting_polity_code))
)
```

China is the reporter that could not go through an area at all: FAOSTAT 351 is the deliberate China
aggregate mapping to no polity, so its ten aliases target the `CHN` chain directly — which the alias
map permits, because it maps labels to polities rather than to areas.

**Keyed by a region taxonomy.** The rest are GLEAM, IPCC, Bouwman, Krausmann and HANPP coefficient
tables. A GLEAM region is no more derivable from a polity than an IPCC region is, so these correctly
carry no polity, and "add a polity column" would be the wrong fix. What they need instead is that
the vocabulary a table is *keyed by* matches the vocabulary it is *joined against* — because a
mismatch yields `NA` coefficients that a fallback then hides, with no error anywhere.

That has already cost something. `residue_feed_fraction.csv` has a column named `region_hanpp`
holding UN M49 sub-regions; it is joined against `regions_full$region_HANPP`; the two share **zero**
values, so every row takes the `Global` default of 0.20 and a ninefold published range (0.05 to
0.45) is discarded. It is `whep#405`, baselined rather than fixed, because which taxonomy each cited
coefficient belongs to is a scientific judgement.

```{r}
feed <- whep_coef_table("residue_feed_fraction")
named <- setdiff(unique(feed$region_hanpp), "Global")
c(
  file_values = length(named),
  in_region_HANPP = length(intersect(named, regions_full$region_HANPP)),
  in_region_UN_sub = length(intersect(named, regions_full$region_UN_sub))
)
```

## Where the gaps are

Stated rather than hidden, because each is a real limit on what the data can say:

- **Rest-of-world folding.** Areas with no data of their own fold into a `ROW-*` polity, which is
  what FABIO does. Areas that *do* report data are excluded from the fold, decided by the
  `observed_rows` column of the published alias map.
- **Pre-1961 back-casting** needs a LUH2 land series per area. The French overseas departments
  have none — LUH2 folds *départements* into France — so their earlier production is not
  back-cast rather than being estimated from an aggregate.
- **IPCC regions** come from the GLEAM hierarchy, which does not cover 59 of the areas carrying
  an ISO3. Those use the `Global` coefficient, which covers every category.
- **Reporting buckets** (`RLAM-1850-2013` and the other regional "Other" rows) end before the
  data does, so their most recent years resolve to nothing.
- **Most of the crosswalk is unreachable, by design.** The crosswalk deliberately carries whole
  polity chains, while FAOSTAT-sourced resolution can only deliver the 1961-onward part of each.
  A mapping existing in the crosswalk is therefore not evidence that any query returns it:

```{r}
pairs <- unique(polity_area_crosswalk[
  !is.na(polity_area_crosswalk$area_code) & !is.na(polity_area_crosswalk$polity_code),
  c("area_code", "polity_code", "polity_end_year")
])
grid <- expand.grid(
  area_code = sort(unique(pairs$area_code)),
  year = 1961:2024
)
resolved <- add_polity_code(grid)
reachable <- unique(paste(resolved$area_code, resolved$polity_code))
unreachable <- pairs[!paste(pairs$area_code, pairs$polity_code) %in% reachable, ]
c(
  pairs = nrow(pairs),
  unreachable = nrow(unreachable),
  ended_by_1961 = sum(unreachable$polity_end_year <= 1961L)
)
```

  Almost every unreachable pair belongs to a polity that ended at or before 1961, which the clamp
  makes unselectable.

```{r}
# The one unreachable pair that does NOT end before 1961 — area 15 maps to two overlapping
# Belgium-Luxembourg rows and the resolver only ever picks the first.
add_polity_code(data.frame(area_code = 15L, year = c(1930L, 1970L)))$polity_code
```
