---
title: "How areas resolve to polities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How areas resolve to polities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(whep)
```

Every area, country and region identifier in this package resolves to a **polity** — a
territorial-economic unit from the `whep-polities` database, coded `PREFIX-startyear-endyear`.
This vignette documents how that resolution works, which columns are safe to join on, and
where the honest gaps are.

## What the package embeds, and what it verifies

`whep-polities` publishes four files. This package embeds a copy of each and checks it has not
drifted:

| upstream file | embedded as | what is verified |
|---|---|---|
| `polities_database.gpkg` | `polities` | row-by-row on eight identity fields |
| `faostat_area_polity_map.csv` | `polity_area_crosswalk` | no area conflicts on overlapping years |
| `label_alias_map.csv` | `polity_label_aliases` | every alias targets a polity this package carries |
| `polities_manifest.json` | — | identity fields, dead codes, polygon gaps |

Those comparisons need the upstream repository, which is private, so they skip on CI. What runs
everywhere is the comparison **between the two embedded objects**: `polities` and
`polity_area_crosswalk` are built by separate steps from the same source, so disagreement on any
identity field means one was regenerated and the other was not.

## Resolving an area

```{r}
add_polity_code(data.frame(area_code = c(185L, 51L), year = c(2000L, 1960L)))[
  , c("area_code", "year", "polity_code")
]
```

Two conventions matter:

- **`end_year` is exclusive.** Area 185 in 2014 resolves to `RUS-2014-2025`, not `RUS-1991-2014`.
- **Years before 1961 match the polity active in 1961.** Not a convenience: WHEP's pre-1961
  series are *back-cast onto 1961 territory* rather than reported under their own year's borders,
  so the 1961-active polity is the correct target and matching by data year would attribute
  1961-border figures to an entity with different borders. On the real production build this
  covers 1,978,400 rows — 31% of the dataset — whose polity span excludes the row's own year.
  Pass `backcast_anchor = -Inf` to match strictly by data year instead.

A resolved row carries both a code and a numeric key, and they answer different questions:

- `reporting_polity_code` — the polity. Join this to `polities`.
- `polity_area_code` — the numeric area key the matrix workflows use. Most builds emit this
  **as** their `area_code`, so it, not the polity code, determines how rows aggregate.

That distinction is load-bearing. Eleven areas that FABIO folds into rest-of-world report data of
their own, and correcting only their `polity_code` left them still summed into area 999 by the
numeric key.

## Resolving a label

`resolve_polity_label()` tries two routes in order: the alias map first, then the polity's
own name. Aliases are source-scoped and year-ranged, so naming both gets the curated answer:

```{r}
resolve_polity_label(
  c("Ivory Coast", "Türkiye", "Eswatini"),
  source = "faostat",
  year = 2000L
)
```

```{r}
# Without a year, a year-ranged alias cannot match, and no polity is NAMED
# "Ivory Coast" — the database calls it "Côte d'Ivoire".
resolve_polity_label("Ivory Coast", source = "faostat")
```

The name route exists because the database's own names were unreachable through the alias
map. `resolve_polity_label("Netherlands")` returned `NA` while `polities` carried a polity
named exactly that, so a caller spelling an area the ordinary way got nothing:

```{r}
resolve_polity_label(
  c("Netherlands", "Tanzania", "United Kingdom"),
  year = 2000L
)
```

Two guards decide when a name may answer, and both matter more than they look.

A name resolves only when **exactly one** polity carries it in the year asked about. 52 pairs
of polities share a normalised name and overlap in years: `polity_name` carries parenthesised
qualifiers like `"Zimbabwe (1900-1953)"` which normalisation drops, and nested periodisations
put two live candidates in one year. Picking by row order would invent an answer.

```{r}
# Two Peru periods cover 1865, so the name alone cannot say which.
resolve_polity_label(c("Peru", "Netherlands"), year = 1865L)
```

And a curated rule covering that year outranks the name, whatever its source. Every alias for
`"djibouti"` routes to the French Somaliland family, so the name route is not allowed to
answer `DJI-1886-2025` over it:

```{r}
resolve_polity_label("djibouti", year = 1961L)
resolve_polity_label("burundi", source = "iia", year = 1930L)
```

Outside its own span a year-ranged rule is silent rather than contradicted, so the name
answers there:

```{r}
# natal's only alias covers 1910-1957; Morocco's start at 1961.
resolve_polity_label(c("natal", "Morocco"), year = 1900L)
```

## Three columns look like an area name; one is safe

`regions_full` carries `name`, `FAOSTAT_name` and `polity_name`. Only one matches the crosswalk:

```{r}
cw <- unique(as.data.frame(polity_area_crosswalk)[, c("area_code", "area_name")])
rf <- as.data.frame(regions_full)[, c("code", "name", "FAOSTAT_name", "polity_name")]
both <- merge(cw[!is.na(cw$area_code), ], rf, by.x = "area_code", by.y = "code")
vapply(
  c("name", "FAOSTAT_name", "polity_name"),
  function(col) sum(both$area_name != both[[col]], na.rm = TRUE),
  integer(1)
)
```

`FAOSTAT_name` is canonical. `name` carries short forms (Turkey, Czech Republic, Ivory Coast) and
`polity_name` carries the aggregate label for folded areas (Bermuda reads "Latin America Other").
A lot of code joins on `c("year", "area", ...)`, and using either non-canonical column loses rows
silently — an unmatched join yields `NA`, not an error. Two inputs had exactly that defect:
`gdp-population` lost 6.4% of its rows and the fodder bridge 13.3%.

## A source that spells an area its own way

`add_area_code()` matched the crosswalk's `area_name` exactly, and the crosswalk carries
FAOSTAT long forms. Any source using an ordinary short form got `NA` even though the code was
sitting there — `"Netherlands"` against `"Netherlands (Kingdom of the)"`, `"Tanzania"` against
`"United Republic of Tanzania"`. Names the exact join cannot match now route through this
database instead, so both spellings reach the same code:

```{r}
add_area_code(
  tibble::tibble(
    area = c("Netherlands", "Netherlands (Kingdom of the)", "Dummy Country"),
    year = 2000L
  ),
  name_column = "area"
)
```

The fallback fills `NA` only, so the crosswalk stays authoritative wherever it has an answer
and this cannot move data that already matched. It is year-aware because the answer depends on
the year: `"Czech Republic"` is Czechia (area 167) from 1993 and Czechoslovakia (area 51)
before it.

`get_primary_residues()` is the one builder that resolves its areas by name, and it measures
what this was worth. Of 475,688 residue rows, 44,985 (9.5%) carried no area code across 14
labels. Routing unmatched names through the polities database recovered 34,963 of them;
rescoping four aliases upstream from `source=faostat` to blanket recovered another 9,822. What
remains is 200 rows — 0.04% — all Tanzania 1961-1964, held up by the `TAN-1922-1964`
code/column disagreement that whep-polities baselines rather than guesses at.

## Datasets that are not keyed by a FAOSTAT area

Everything above is about `area_code`. But a census of the package's data — the lazy-loaded
datasets *and* the internal ones, since an `ls()` over the namespace alone misses lazy data and
would report a single table — finds 21 area-keyed tables with no polity column. They split cleanly
in two, and the split is the whole point: one half should carry polities and now does, the other
half should not.

**Keyed by a territory.** Each resolves through the published alias map, under a source name, with
year-scoped aliases maintained upstream rather than name-matched here:

```{r}
resolve_share <- function(labels, source, year) {
  round(100 * mean(!is.na(resolve_polity_label(labels, source = source, year = year))), 1)
}
lass <- unique(lassaletta_grassland_share[, c("Country", "year")])
by_year <- tapply(
  seq_len(nrow(lass)),
  lass$year,
  function(i) {
    !is.na(resolve_polity_label(
      as.character(lass$Country[i]),
      source = "lassaletta-grassland-share",
      year = lass$year[i][[1]]
    ))
  }
)
c(
  lassaletta_rows = nrow(lass),
  lassaletta_pct = round(100 * mean(unlist(by_year)), 1),
  crops_manure_pct = resolve_share(
    unique(stats::na.omit(crops_manure_n$ISO)),
    "crops-manure-n",
    2000L
  ),
  mueller_pct = resolve_share(
    unique(stats::na.omit(mueller_synthetic_n$iso3c)),
    "mueller-synthetic-n",
    2000L
  )
)
```

Mueller and `crops_manure_n` are complete, and Lassaletta is complete in the only sense available
to it: every one of its remaining unresolved country-years is a **source anachronism** — a label
carried outside the period its entity existed. `South Sudan` is listed from 1961 though it did not
exist until 2011; `FSU` and `Yugoslav SFR` run past their dissolutions; `Czechoslovakia` reaches
2009 though it dissolved in 1993; `Belgium-Luxemburg` continues past the 1999 split. Each resolves
correctly for the years its entity *did* exist, so resolving the tails would be wrong rather than
better.

`expand_trade_sources()` is the same shape, and complete: its rows are already one year each, so
the reporter and the year settle every row.

```{r}
src <- utils::read.csv(
  system.file("extdata", "trade_sources.csv", package = "whep"),
  check.names = FALSE
)
out <- expand_trade_sources(src)
c(
  rows = nrow(out),
  with_polity = sum(!is.na(out$reporting_polity_code)),
  distinct_polities = length(unique(out$reporting_polity_code))
)
```

China is the reporter that could not go through an area at all: FAOSTAT 351 is the deliberate China
aggregate mapping to no polity, so its ten aliases target the `CHN` chain directly — which the alias
map permits, because it maps labels to polities rather than to areas.

**Keyed by a region taxonomy.** The rest are GLEAM, IPCC, Bouwman, Krausmann and HANPP coefficient
tables. A GLEAM region is no more derivable from a polity than an IPCC region is, so these correctly
carry no polity, and "add a polity column" would be the wrong fix. What they need instead is that
the vocabulary a table is *keyed by* matches the vocabulary it is *joined against* — because a
mismatch yields `NA` coefficients that a fallback then hides, with no error anywhere.

That has already cost something. `residue_feed_fraction.csv` has a column named `region_hanpp`
holding UN M49 sub-regions; it is joined against `regions_full$region_HANPP`; the two share **zero**
values, so every row takes the `Global` default of 0.20 and a ninefold published range (0.05 to
0.45) is discarded. It is `whep#405`, baselined rather than fixed, because which taxonomy each cited
coefficient belongs to is a scientific judgement.

```{r}
feed <- whep_coef_table("residue_feed_fraction")
named <- setdiff(unique(feed$region_hanpp), "Global")
c(
  file_values = length(named),
  in_region_HANPP = length(intersect(named, regions_full$region_HANPP)),
  in_region_UN_sub = length(intersect(named, regions_full$region_UN_sub))
)
```

## What a clean build says

A build that resolves every row still emits one warning, and knowing which one is the difference between reading
output and trusting it. On `build_primary_production(1990, 1993)` chained into
`build_commodity_balances(p, 1990, 1993)` — 165,417 and 390,020 rows, **zero `NA` across all four polity columns in
both** (checked column by column, not in aggregate):

```
LUH2 territories with no FAOSTAT area code, dropping: JEY, GGY, IMN, BLM, ALA, and SXM.
Each has a polity upstream but no area code to aggregate through; see whep#407.
```

A **full-range** build says more, and the difference is worth stating because a narrow window is not a licence to
claim a clean build. `get_wide_cbs()` at its default range completes with **~2.766 million rows across 195 areas and
zero `NA` in all four polity columns**. The row count is given loosely on purpose: the build is
not reproducible run to run, alternating between 2,768,578 and 2,764,471 on an unchanged tree
(whep#420), so any claim resting on a few thousand rows is inside its own noise, and emits seven warnings:

| warning | status |
|---|---|
| LUH2 territories with no FAOSTAT area code (six, twice — once per land read) | known, whep#407 |
| Historical extension: **5** areas have no LUH2 land match | pinned by identity in `test_luh2_backcast_reach.R` (four in code space; the difference in kind is documented there) |
| **24** crop-residue rows lack any `area_code` | down from 44,985; the remainder is whep#417 |
| **31,642** (key, source) combinations are duplicated at the source cast | pre-existing, now reported — whep#418 |
| `'fun.aggregate' is NULL, but found duplicate row/column combinations` | data.table's own warning for the line above |
| **46,424** CBS rows have supply-use imbalance above 50% | pre-existing, **identical on `main`** |

Two of these deserve more than a row in a table.

The **imbalance count is now compared** rather than caveated. It is a conservation check in `R/conservation.R` and
no test pins the number, so an earlier draft of this section said only that it had not been measured against
`main`. It has been: a full-range `get_wide_cbs()` on `main` gives **46,424** as well, the same figure to the row.
That is not luck — this branch ships the numeric aggregation key identical to `main` for all 267 area codes, and
every column total within 1%.

The **duplicated cast keys are the interesting one**, because they are proof of a live defect rather than a
tolerance. `dcast()` with no `fun.aggregate` answers a duplicate key with `length()`, so those values become row
COUNTS instead of quantities — and data.table says so itself in the warning immediately below. 31,642 combinations
in areas 206 and 999 are affected in the shipped build, and they are affected on `main` too; the difference is that
they were silent there. Fixing it means choosing `sum` or `first` at the cast, which changes published numbers, so
it is whep#418 rather than a decision taken here. Two earlier attempts to fix it upstream instead — by summing the
areas that fold into one bucket — were withdrawn after measurement inflated `food` 266x; see the same issue.

That is the whole inventory. Earlier in this work the full-range build emitted four warnings and three of those
were defects, since fixed:

- **an unknown area code, twice.** 420 "Sub-Saharan Africa" and 265 "China (excluding intra-trade)" are FAOSTAT
  aggregates whose codes sit *below* the `group_code_min` threshold, so the rule of thumb missed them and they
  were reported as codes the project does not know. Upstream now publishes the six such codes explicitly.
- **a duplicate-year warning that was a real cross-territory denominator.** Livestock split shares were computed
  as `value / sum(value)` grouped by `polity_area_code`, and one bucket held two territories — so Eswatini's
  broilers were divided by an Eswatini-plus-rest-of-world total. Two different shares for one key, resolved by row
  order. Rest of World's share was wrong by a factor of 222.
- **a sentinel bundled with real territories.** The LUH2 message above used to include `-99`, LUH2's own
  unassigned marker, which carries **510×** the land of the six territories combined. Reporting them together made
  a property of the source read as a gap in this project's coverage.

The pattern in all three: a warning that fires on every build is read as background. The two that mattered were
distinguishable only by measuring what they dropped — 0.6% of a pin, a 222× share error, and 0.39% of global land
against 0.0008%.

## What is guaranteed, and by which check

Prose is easy to write and hard to trust, so each claim above is enforced somewhere. This is the map, so a
reader can go look rather than take it on faith.

| claim | enforced by |
|---|---|
| the resolver honours the published alias map — all 851 ranged aliases resolve to their own target | `test_alias_map_conformance.R` |
| a name resolves only when one polity carries it in that year, and a year-covering rule outranks it | `test_resolve_polity_label_name_route.R`, by identity |
| every dataset naming a territory is tied to the polities database by something | `test_area_dataset_census.R` — built on the lazydata environment, because package data is not an export |
| folded reporting areas sum into their bucket instead of sitting beside each other in it | `test_polity_bucket_aggregation.R`, mutation-tested |
| duplicate keys never reach the source cast, where values would silently become row counts | a guard in `.select_best_source()`, tested by the same file |
| `regions_full` and `polities_cats` differ only where a modelling choice says they should | `test_polities_cats_vs_regions_full.R` |
| the embedded copies match what upstream publishes | `test_polities_upstream_contract.R` and siblings — 51 assertions, and they need the upstream checkout |

Two of those deserve a caveat rather than a tick. The census asserts each area-carrying dataset is
**referenced** by some test, which is weaker than "its vocabulary agrees with the database" — the per-dataset
tests do that, one at a time. And the upstream contract tests **skip** without the sibling repository, which is
why the pre-push harness runs them explicitly with all five inputs pointed at one checkout: a skipped contract
test and a passing one look identical in a summary line.

## Where the gaps are

Stated rather than hidden, because each is a real limit on what the data can say:

- **Rest-of-world folding.** Areas with no data of their own fold into a `ROW-*` polity, which is
  what FABIO does. Areas that *do* report data are excluded from the fold, decided by the
  `observed_rows` column of the published alias map.
- **Pre-1961 back-casting** needs a LUH2 land series per area. The French overseas departments
  have none — LUH2 folds *départements* into France — so their earlier production is not
  back-cast rather than being estimated from an aggregate.
- **IPCC regions** come from the GLEAM hierarchy, which does not cover 59 of the areas carrying
  an ISO3. Those use the `Global` coefficient, which covers every category.
- **Reporting buckets** (`RLAM-1850-2013` and the other regional "Other" rows) end before the
  data does, so their most recent years resolve to nothing.
- **Most of the crosswalk is unreachable, by design.** The crosswalk deliberately carries whole
  polity chains, while FAOSTAT-sourced resolution can only deliver the 1961-onward part of each.
  A mapping existing in the crosswalk is therefore not evidence that any query returns it:

```{r}
pairs <- unique(polity_area_crosswalk[
  !is.na(polity_area_crosswalk$area_code) & !is.na(polity_area_crosswalk$polity_code),
  c("area_code", "polity_code", "polity_end_year")
])
grid <- expand.grid(
  area_code = sort(unique(pairs$area_code)),
  year = 1961:2024
)
resolved <- add_polity_code(grid)
reachable <- unique(paste(resolved$area_code, resolved$polity_code))
unreachable <- pairs[!paste(pairs$area_code, pairs$polity_code) %in% reachable, ]
c(
  pairs = nrow(pairs),
  unreachable = nrow(unreachable),
  ended_by_1961 = sum(unreachable$polity_end_year <= 1961L)
)
```

  Almost every unreachable pair belongs to a polity that ended at or before 1961, which the clamp
  makes unselectable.

```{r}
# The one unreachable pair that does NOT end before 1961 — area 15 maps to two overlapping
# Belgium-Luxembourg rows and the resolver only ever picks the first.
add_polity_code(data.frame(area_code = 15L, year = c(1930L, 1970L)))$polity_code
```
