---
title: "How areas resolve to polities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How areas resolve to polities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(whep)
```

Every area, country and region identifier in this package resolves to a **polity** — a
territorial-economic unit from the `whep-polities` database, coded `PREFIX-startyear-endyear`.
This vignette documents how that resolution works, which columns are safe to join on, and
where the honest gaps are.

## What the package embeds, and what it verifies

`whep-polities` publishes four files. This package embeds a copy of each and checks it has not
drifted:

| upstream file | embedded as | what is verified |
|---|---|---|
| `polities_database.gpkg` | `polities` | row-by-row on eight identity fields |
| `faostat_area_polity_map.csv` | `polity_area_crosswalk` | no area conflicts on overlapping years |
| `label_alias_map.csv` | `polity_label_aliases` | every alias targets a polity this package carries |
| `polities_manifest.json` | — | identity fields, dead codes, polygon gaps |

Those comparisons need the upstream repository, which is private, so they skip on CI. What runs
everywhere is the comparison **between the two embedded objects**: `polities` and
`polity_area_crosswalk` are built by separate steps from the same source, so disagreement on any
identity field means one was regenerated and the other was not.

## Resolving an area

```{r}
add_polity_code(data.frame(area_code = c(185L, 51L), year = c(2000L, 1960L)))[
  , c("area_code", "year", "polity_code")
]
```

Two conventions matter:

- **`end_year` is exclusive.** Area 185 in 2014 resolves to `RUS-2014-2025`, not `RUS-1991-2014`.
- **Years before 1961 match the polity active in 1961.** Not a convenience: WHEP's pre-1961
  series are *back-cast onto 1961 territory* rather than reported under their own year's borders,
  so the 1961-active polity is the correct target and matching by data year would attribute
  1961-border figures to an entity with different borders. On the real production build this
  covers 1,978,400 rows — 31% of the dataset — whose polity span excludes the row's own year.
  Pass `backcast_anchor = -Inf` to match strictly by data year instead.

A resolved row carries both a code and a numeric key, and they answer different questions:

- `reporting_polity_code` — the polity. Join this to `polities`.
- `polity_area_code` — the numeric area key the matrix workflows use. Most builds emit this
  **as** their `area_code`, so it, not the polity code, determines how rows aggregate.

That distinction is load-bearing, and it is live rather than historical. Sixteen areas that FABIO
folds into rest-of-world report data of their own. Their `polity_code` is corrected to their own
polity, while their `polity_area_code` is still 999 — so **the crosswalk names a polity that the
numbers are not aggregated under**. A reader checking `reporting_polity_code` sees the Faroe
Islands; the values are in `ROW-1850-2023`.

That is deliberate, not an oversight: making the two agree by promoting these areas numerically
inflates global `feed` 13.7×. It is the single most important thing to know before joining on one
of these two columns and assuming the other follows. whep#419.

## Resolving a label

`resolve_polity_label()` tries two routes in order: the alias map first, then the polity's
own name. Aliases are source-scoped and year-ranged, so naming both gets the curated answer:

```{r}
resolve_polity_label(
  c("Ivory Coast", "Türkiye", "Eswatini"),
  source = "faostat",
  year = 2000L
)
```

```{r}
# Without a year, a year-ranged alias cannot match, and no polity is NAMED
# "Ivory Coast" — the database calls it "Côte d'Ivoire".
resolve_polity_label("Ivory Coast", source = "faostat")
```

The name route exists because the database's own names were unreachable through the alias
map. `resolve_polity_label("Netherlands")` returned `NA` while `polities` carried a polity
named exactly that, so a caller spelling an area the ordinary way got nothing:

```{r}
resolve_polity_label(
  c("Netherlands", "Tanzania", "United Kingdom"),
  year = 2000L
)
```

Two guards decide when a name may answer, and both matter more than they look.

A name resolves only when **exactly one** polity carries it in the year asked about. 52 pairs
of polities share a normalised name and overlap in years: `polity_name` carries parenthesised
qualifiers like `"Zimbabwe (1900-1953)"` which normalisation drops, and nested periodisations
put two live candidates in one year. Picking by row order would invent an answer.

```{r}
# Two Peru periods cover 1865, so the name alone cannot say which.
resolve_polity_label(c("Peru", "Netherlands"), year = 1865L)
```

And a curated rule covering that year outranks the name, whatever its source. Every alias for
`"djibouti"` routes to the French Somaliland family, so the name route is not allowed to
answer `DJI-1886-2025` over it:

```{r}
resolve_polity_label("djibouti", year = 1961L)
resolve_polity_label("burundi", source = "iia", year = 1930L)
```

Outside its own span a year-ranged rule is silent rather than contradicted, so the name
answers there:

```{r}
# natal's only alias covers 1910-1957; Morocco's start at 1961.
resolve_polity_label(c("natal", "Morocco"), year = 1900L)
```

## Three columns look like an area name; one is safe

`regions_full` carries `name`, `FAOSTAT_name` and `polity_name`. Only one matches the crosswalk:

```{r}
cw <- unique(as.data.frame(polity_area_crosswalk)[, c("area_code", "area_name")])
rf <- as.data.frame(regions_full)[, c("code", "name", "FAOSTAT_name", "polity_name")]
both <- merge(cw[!is.na(cw$area_code), ], rf, by.x = "area_code", by.y = "code")
vapply(
  c("name", "FAOSTAT_name", "polity_name"),
  function(col) sum(both$area_name != both[[col]], na.rm = TRUE),
  integer(1)
)
```

`FAOSTAT_name` is canonical. `name` carries short forms (Turkey, Czech Republic, Ivory Coast) and
`polity_name` carries the aggregate label for folded areas (Bermuda reads "Latin America Other").
A lot of code joins on `c("year", "area", ...)`, and using either non-canonical column loses rows
silently — an unmatched join yields `NA`, not an error. Two inputs had exactly that defect:
`gdp-population` lost 6.4% of its rows and the fodder bridge 13.3%.

## A source that spells an area its own way

`add_area_code()` matched the crosswalk's `area_name` exactly, and the crosswalk carries
FAOSTAT long forms. Any source using an ordinary short form got `NA` even though the code was
sitting there — `"Netherlands"` against `"Netherlands (Kingdom of the)"`, `"Tanzania"` against
`"United Republic of Tanzania"`. Names the exact join cannot match now route through this
database instead, so both spellings reach the same code:

```{r}
add_area_code(
  tibble::tibble(
    area = c("Netherlands", "Netherlands (Kingdom of the)", "Dummy Country"),
    year = 2000L
  ),
  name_column = "area"
)
```

The fallback fills `NA` only, so the crosswalk stays authoritative wherever it has an answer
and this cannot move data that already matched. It is year-aware because the answer depends on
the year: `"Czech Republic"` is Czechia (area 167) from 1993 and Czechoslovakia (area 51)
before it.

`get_primary_residues()` is the one builder that resolves its areas by name, and it measures
what this was worth. Of 475,688 residue rows, 44,985 (9.5%) carried no area code across 14
labels. Routing unmatched names through the polities database recovered 34,963 of them;
rescoping four aliases upstream from `source=faostat` to blanket recovered another 9,822. What
remains is 200 rows — 0.04% — all Tanzania 1961-1964, held up by the `TAN-1922-1964`
code/column disagreement that whep-polities baselines rather than guesses at.

## Datasets that are not keyed by a FAOSTAT area

Everything above is about `area_code`. But a census of the package's data — the lazy-loaded
datasets *and* the internal ones, since an `ls()` over the namespace alone misses lazy data and
would report a single table — finds 21 area-keyed tables with no polity column. They split cleanly
in two, and the split is the whole point: one half should carry polities and now does, the other
half should not.

**Keyed by a territory.** Each resolves through the published alias map, under a source name, with
year-scoped aliases maintained upstream rather than name-matched here:

```{r}
resolve_share <- function(labels, source, year) {
  round(100 * mean(!is.na(resolve_polity_label(labels, source = source, year = year))), 1)
}
lass <- unique(lassaletta_grassland_share[, c("Country", "year")])
by_year <- tapply(
  seq_len(nrow(lass)),
  lass$year,
  function(i) {
    !is.na(resolve_polity_label(
      as.character(lass$Country[i]),
      source = "lassaletta-grassland-share",
      year = lass$year[i][[1]]
    ))
  }
)
c(
  lassaletta_rows = nrow(lass),
  lassaletta_pct = round(100 * mean(unlist(by_year)), 1),
  crops_manure_pct = resolve_share(
    unique(stats::na.omit(crops_manure_n$ISO)),
    "crops-manure-n",
    2000L
  ),
  mueller_pct = resolve_share(
    unique(stats::na.omit(mueller_synthetic_n$iso3c)),
    "mueller-synthetic-n",
    2000L
  )
)
```

Mueller and `crops_manure_n` are complete, and Lassaletta is complete in the only sense available
to it: every one of its remaining unresolved country-years is a **source anachronism** — a label
carried outside the period its entity existed. `South Sudan` is listed from 1961 though it did not
exist until 2011; `FSU` and `Yugoslav SFR` run past their dissolutions; `Czechoslovakia` reaches
2009 though it dissolved in 1993; `Belgium-Luxemburg` continues past the 1999 split. Each resolves
correctly for the years its entity *did* exist, so resolving the tails would be wrong rather than
better.

`expand_trade_sources()` is the same shape, and complete: its rows are already one year each, so
the reporter and the year settle every row.

```{r}
src <- utils::read.csv(
  system.file("extdata", "trade_sources.csv", package = "whep"),
  check.names = FALSE
)
out <- expand_trade_sources(src)
c(
  rows = nrow(out),
  with_polity = sum(!is.na(out$reporting_polity_code)),
  distinct_polities = length(unique(out$reporting_polity_code))
)
```

China is the reporter that could not go through an area at all: FAOSTAT 351 is the deliberate China
aggregate mapping to no polity, so its ten aliases target the `CHN` chain directly — which the alias
map permits, because it maps labels to polities rather than to areas.

**Keyed by a region taxonomy.** The rest are GLEAM, IPCC, Bouwman, Krausmann and HANPP coefficient
tables. A GLEAM region is no more derivable from a polity than an IPCC region is, so these correctly
carry no polity, and "add a polity column" would be the wrong fix. What they need instead is that
the vocabulary a table is *keyed by* matches the vocabulary it is *joined against* — because a
mismatch yields `NA` coefficients that a fallback then hides, with no error anywhere.

That has already cost something. `residue_feed_fraction.csv` has a column named `region_hanpp`
holding UN M49 sub-regions; it is joined against `regions_full$region_HANPP`; the two share **zero**
values, so every row takes the `Global` default of 0.20 and a ninefold published range (0.05 to
0.45) is discarded. It is `whep#405`, baselined rather than fixed, because which taxonomy each cited
coefficient belongs to is a scientific judgement.

```{r}
feed <- whep_coef_table("residue_feed_fraction")
named <- setdiff(unique(feed$region_hanpp), "Global")
c(
  file_values = length(named),
  in_region_HANPP = length(intersect(named, regions_full$region_HANPP)),
  in_region_UN_sub = length(intersect(named, regions_full$region_UN_sub))
)
```

## What a clean build says

A build that resolves every row still emits one warning, and knowing which one is the difference between reading
output and trusting it. On `build_primary_production(1990, 1993)` chained into
`build_commodity_balances(p, 1990, 1993)` — 165,417 and 390,020 rows, **zero `NA` across all four polity columns in
both** (checked column by column, not in aggregate):

```
LUH2 territories with no FAOSTAT area code, dropping: JEY, GGY, IMN, BLM, ALA, and SXM.
Each has a polity upstream but no area code to aggregate through; see whep#407.
```

A **full-range** build says more, and the difference is worth stating because a narrow window is not a licence to
claim a clean build. `get_wide_cbs()` at its default range completes with **~2.00 million rows across 195 areas and
zero `NA` in all four polity columns**. The row count is given loosely on purpose: the build is
not reproducible run to run — 1,999,609 twice in one session and 1,997,944 in another
(whep#420) — so any claim resting on a few thousand rows is inside its own noise, and emits six warnings:

The count fell from ~2.766M when whep#425 was fixed, and the drop is the fix working rather than
data lost: row counts are never zero, so the `value != 0` filter had nothing to remove until
quantities were restored. About 390,000 genuine zeros now leave.

| warning | status |
|---|---|
| LUH2 territories with no FAOSTAT area code (six, twice — once per land read) | known, whep#407 |
| Historical extension: **5** areas have no LUH2 land match | pinned by identity in `test_luh2_backcast_reach.R` (four in code space; the difference in kind is documented there) |
| **200** crop-residue rows resolve to no area (label `"Tanzania"`, years 1961-1964) | added by this branch so the loss is visible rather than silent; whep#417 |
| **24** crop-residue rows lack any `area_code` | down from 44,985; the remainder is whep#417 |
| **47,514** CBS rows have supply-use imbalance above 50% | pre-existing, **identical on `main` to the row**, and on branch-code-with-main-data too |

Counted rather than tallied by hand: a full-range build with every warning captured emits exactly
six, which is the five rows above with the LUH2 one appearing twice. The same capture confirms
what the repurposed cast guard is supposed to do — **no** warning matching `fold no other`,
`fun.aggregate` or `row count` is emitted, so the guard is silent on a healthy build and
data.table's own duplicate-combination warning is gone.

Two of these deserve more than a row in a table.

The **imbalance count is now compared** rather than caveated. It is a conservation check in `R/conservation.R` and
no test pins the number, so an earlier draft of this section said only that it had not been measured against
`main`. It has been: a full-range `get_wide_cbs()` on `main` gives **47,514** as well, the same figure to the row,
and so does branch-code-with-main-data. Three independent builds, one number.

That invariance is worth separating from a claim it was previously bundled with. The imbalance count and the
numeric aggregation key (identical for all 267 area codes) are neutral. **Column totals are not** — measured on the
fixed build, `feed` moves +7.0%, `other_uses` +3.1%, `production` +2.0% and `domestic_supply` +1.9% against `main`,
concentrated in FABIO's rest-of-world bucket. An earlier version of this vignette said every column total was
within 1%; that was measured while whep#425 made every value a row count, and row counts compare group sizes
rather than quantities, so it could not have detected what it was taken to rule out. The movement is unexplained
and tracked in whep#444.

The **duplicated cast keys are gone from this list**, because whep#425 is fixed (whep#429). `dcast()` with no
`fun.aggregate` answered a duplicate key with `length()` and applied that to every cell, so all values became row
COUNTS. The cast now supplies `fun.aggregate` and sums them: 31,642 duplicated `(key, source)` combinations still
arrive, in areas 206 and 999, because they are one reporting bucket's folded members and that is upstream of the
cast — but summing is what a bucket's total means.

The warning this vignette used to describe has been **repurposed rather than removed**. It said casting would
replace values with row counts, which is now false, and a loud warning about correct behaviour is noise that gets
muffled. It now fires only when duplicates appear in a bucket that folds **no** other area, which would mean rows
are colliding for a reason folding does not explain. The fold set is derived from `polity_area_crosswalk` at build
time — 206, 238 and 999 today — so a fourth fold appearing surfaces instead of being hardcoded away.

That is the whole inventory. Earlier in this work the full-range build emitted four warnings and three of those
were defects, since fixed:

- **an unknown area code, twice.** 420 "Sub-Saharan Africa" and 265 "China (excluding intra-trade)" are FAOSTAT
  aggregates whose codes sit *below* the `group_code_min` threshold, so the rule of thumb missed them and they
  were reported as codes the project does not know. Upstream now publishes the six such codes explicitly.
- **a duplicate-year warning that was a real cross-territory denominator.** Livestock split shares were computed
  as `value / sum(value)` grouped by `polity_area_code`, and one bucket held two territories — so Eswatini's
  broilers were divided by an Eswatini-plus-rest-of-world total. Two different shares for one key, resolved by row
  order. Rest of World's share was wrong by a factor of 222.
- **a sentinel bundled with real territories.** The LUH2 message above used to include `-99`, LUH2's own
  unassigned marker, which carries **510×** the land of the six territories combined. Reporting them together made
  a property of the source read as a gap in this project's coverage.

The pattern in all three: a warning that fires on every build is read as background. The two that mattered were
distinguishable only by measuring what they dropped — 0.6% of a pin, a 222× share error, and 0.39% of global land
against 0.0008%.

## What is guaranteed, and by which check

Prose is easy to write and hard to trust, so each claim above is enforced somewhere. This is the map, so a
reader can go look rather than take it on faith.

| claim | enforced by |
|---|---|
| the resolver honours the published alias map — all 851 ranged aliases resolve to their own target | `test_alias_map_conformance.R` |
| a name resolves only when one polity carries it in that year, and a year-covering rule outranks it | `test_resolve_polity_label_name_route.R`, by identity |
| every dataset naming a territory is tied to the polities database by something | `test_area_dataset_census.R` — built on the lazydata environment, because package data is not an export |
| territory-periods sharing a bucket are kept apart, because `area` is a join key and the bucket's total belongs at the cast (whep#425) | `test_polity_bucket_aggregation.R`, mutation-tested |
| the aggregator's `area` label agrees with what the residues path emits, since `area` is a join key | the same file — two vocabularies for it once dropped 702,166 rows silently |
| duplicates reaching the source cast are **summed**, and unexplained ones are reported | whep#425 fixed the cast (whep#429). The guard in `.select_best_source()` now fires only for duplicates in a bucket that folds no other area, with the fold set derived from the crosswalk |
| the numeric aggregation key honours FABIO's fold, and both published tables state it identically | `test_polity_area_code_not_folded.R` — promoting these areas inflated global `feed` 13.7× (#419) |
| `regions_full` and `polities_cats` differ only where a modelling choice says they should | `test_polities_cats_vs_regions_full.R` |
| the embedded copies match what upstream publishes | `test_polities_upstream_contract.R` and siblings — 51 assertions, and they need the upstream checkout |
| the embedded data corresponds to a **recorded** upstream state, checkable without that checkout | `test_upstream_stamp.R` against `inst/extdata/upstream_stamp.json`; 9 of its 12 assertions run on CI |
| **no polity is named in package logic** — identity always resolves through the database | `check_like_ci.R` gate "no polity code in package logic", mutation-tested. Zero literals across 122 files; the one match in `R/` is a roxygen line explaining that a family prefix is not a code. Tests are excluded deliberately: a literal there is a fact pinned by identity, in logic it is a bypass that survives a polity being retired upstream |
| every field the contract publishes is actually consumed | audited field by field: `identity_fields` drives the comparison, `live_polity_codes` + `dead_polity_codes` check the code set, `faostat_unmapped_areas` feeds `data-raw/constants.R`, `local_iso3_codes` is cross-checked in `test_iso3_polity_coverage.R`. No published field is dead weight, and nothing the consumer needs is inferred where upstream could state it |
| no published column total drifts unnoticed | `inst/scripts/compare_cbs_totals.R` against a committed baseline — the gap that let two changes inflate values up to 266× while the suite stayed green |
| a dataset or extdata change is real rather than serialisation churn | `inst/scripts/compare_datasets_to_ref.R`, mutation-tested both ways — 27 of 33 changed `.rda` files here are byte-identical in content |
| ISO3 codes used by these tables either resolve to a polity or are a documented exemption | `test_iso3_polity_coverage.R`, pinned by identity in four classes |
| the eight primary builders attach and document polity columns | `check_like_ci.R`'s documentation gate plus `test_builder_polity_columns_populated.R` — **22 further area-keyed exported outputs carry the polity-derived `area_code` but not the columns**, which is whep#424 |

One property is worth stating separately because it is the strongest form of "interconnected" available:
**the committed data is exactly reproducible from the published contract.** Re-running
`data-raw/table_mappings.R` and `data-raw/harmonization_tables.R` against `whep-polities`'s
`data/final/` leaves `git status` empty — not merely equivalent content, but identical bytes across all 34
datasets, the two harmonization CSVs, `R/sysdata.rda` and the upstream stamp. So a reviewer can regenerate
rather than trust, and a drifted dataset is a diff rather than an argument.

Two of those deserve a caveat rather than a tick. The census asserts each area-carrying dataset is
**referenced** by some test, which is weaker than "its vocabulary agrees with the database" — the per-dataset
tests do that, one at a time. And the upstream contract tests **skip** without the sibling repository, which is
why the pre-push harness runs them explicitly with all five inputs pointed at one checkout: a skipped contract
test and a passing one look identical in a summary line.

## One gap that looks like five

Four polities account for most of the separate-looking gaps below, and finding that out took
five independent discoveries, so it is stated once here rather than left to be rediscovered:

| polity | area |
|---|---|
| `F51-1947-1993` | Czechoslovakia |
| `F228-1945-1991` | USSR |
| `F248-1991-1992` | Yugoslav SFR |
| `SCG-1992-2006` | Serbia and Montenegro |

They report data, they carry **11.88% of production value at 1961**, and every present-day
reference table excludes them — correctly, because they no longer exist:

- no LUH2 land series, so their pre-1961 production is not back-cast (`test_luh2_backcast_reach.R`)
- no row in GLEAM's country table, so the energy extension cannot group them
  (`test_gleam_hierarchy_coverage.R`)
- no assignment in six region schemes — UN sub-region, ILO 1–3, IEA, IPCC
  (`test_region_classifications.R`)
- no `iso3_code` in the polities database for three of the four, while the package's own tables
  reference `CSK` and `SUN` — and the database is inconsistent here, since Zaire's twelve polities
  all carry their successor's `COD` (lbm364dl/whep-polities#55)

The single cause is that a dissolved state has no present-day identifier to join on, and each
downstream table solves that by omission. So the practical rule for a reader: **any aggregation
that groups by a present-day classification silently loses roughly a tenth of the 1961 record**,
and it will not warn. The tests above pin the four by identity so that a *live* country joining
that set surfaces as a change rather than hiding inside a count.

## Where the gaps are

Stated rather than hidden, because each is a real limit on what the data can say:

- **Rest-of-world folding, and it is not resolved.** Areas FABIO folds into rest-of-world take a
  `ROW-*` polity. Sixteen of them report data of their own, so that data is attributed to
  `ROW-1850-2023` — the Faroe Islands' 2,458 production rows, Palestine's 9,606. At the *polity*
  level they are excluded from the fold, decided by the `observed_rows` column of the published
  alias map; at the **numeric** level (`polity_area_code`), which is what the builds aggregate on,
  they still fold. That split is deliberate and measured: promoting them numerically inflates
  global `feed` **13.7×**, with the whole increase landing on one area. The cost of not promoting
  them is measured too — 1,389,871 t of feed demand dropped for want of a Bouwman region on RoW.
  See whep#419; neither figure is a rounding error and the choice is a modelling decision.
- **Pre-1961 back-casting** needs a LUH2 land series per area, and the four dissolved federations
  have none (below).
- **IPCC regions** come from the GLEAM hierarchy, which does not cover 59 of the 263 areas that
  carry both an ISO3 and an `area_code` (65 of 269 if areas with no `area_code` are counted).
  Those use the `Global` coefficient, which covers every category.
- **Reporting buckets** (`RLAM-1850-2013` and the other regional "Other" rows) end before the
  data does, so their most recent years resolve to nothing.
- **Most of the crosswalk is unreachable, by design.** The crosswalk deliberately carries whole
  polity chains, while FAOSTAT-sourced resolution can only deliver the 1961-onward part of each.
  A mapping existing in the crosswalk is therefore not evidence that any query returns it:

```{r}
pairs <- unique(polity_area_crosswalk[
  !is.na(polity_area_crosswalk$area_code) & !is.na(polity_area_crosswalk$polity_code),
  c("area_code", "polity_code", "polity_end_year")
])
grid <- expand.grid(
  area_code = sort(unique(pairs$area_code)),
  year = 1961:2024
)
resolved <- add_polity_code(grid)
reachable <- unique(paste(resolved$area_code, resolved$polity_code))
unreachable <- pairs[!paste(pairs$area_code, pairs$polity_code) %in% reachable, ]
c(
  pairs = nrow(pairs),
  unreachable = nrow(unreachable),
  ended_by_1961 = sum(unreachable$polity_end_year <= 1961L)
)
```

  Almost every unreachable pair belongs to a polity that ended at or before 1961, which the clamp
  makes unselectable.

```{r}
# The one unreachable pair that does NOT end before 1961 — area 15 maps to two overlapping
# Belgium-Luxembourg rows and the resolver only ever picks the first.
add_polity_code(data.frame(area_code = 15L, year = c(1930L, 1970L)))$polity_code
```
