---
title: "Trade sources year coverage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trade sources year coverage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Plots require ggplot2 (Suggests); skip them if it is not installed so the
# vignette still builds offline on minimal setups.
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r setup}
library(whep)
```

First we read the bundled snapshot of the trade sources table and build a
dataframe where each row accounts for one year. The snapshot ships with the
package so the vignette builds without any network access. It mirrors the
`Final_Sources_Trade` tab of the project's trade sources spreadsheet.
```{r}
trade_sources_path <- system.file(
  "extdata",
  "trade_sources.csv",
  package = "whep"
)

expanded_trade_sources <-
  trade_sources_path |>
  utils::read.csv(check.names = FALSE, na.strings = c("", "NA")) |>
  tibble::as_tibble() |>
  expand_trade_sources()
```

Now we build some plots.

Plot showing years covered by `expanded_trade_sources`:
```{r, eval = has_ggplot2, fig.alt="Plot showing years covered by expanded_trade_sources"}
ggplot2::ggplot(
  expanded_trade_sources,
  ggplot2::aes(y = Trade, x = Year, fill = "lightblue")
) +
  ggplot2::geom_tile(alpha = .8) +
  ggplot2::theme_dark() +
  ggplot2::labs(title = "Source Availability by Country") +
  ggplot2::scale_fill_identity() +
  ggplot2::facet_wrap(~Reporter, ncol = 1)
```

Plot showing by years, colored by `expanded_trade_sources`:
```{r, eval = has_ggplot2, fig.alt="Plot showing by years, colored by expanded_trade_sources"}
ggplot2::ggplot(
  expanded_trade_sources,
  ggplot2::aes(y = Trade, x = Year, fill = Name, alpha = .8)
) +
  ggplot2::geom_tile() +
  # theme_minimal() +
  ggplot2::theme(legend.position = "None") +
  ggplot2::labs(title = "Source Availability by Country") +
  ggplot2::facet_grid(ImpExp ~ Reporter, scales = "free", space = "free_x")
```

Plot showing years covered by `expanded_trade_sources`, colored by quality:
```{r, eval = has_ggplot2, fig.alt="Plot showing years covered by expanded_trade_sources, colored by quality"}
ggplot2::ggplot(
  expanded_trade_sources, ggplot2::aes(y = Trade, x = Year, alpha = .8)
) +
  ggplot2::geom_tile(ggplot2::aes(fill = factor(Quality))) +
  ggplot2::theme_dark() +
  ggplot2::labs(title = "Source Availability by Country", fill = "Quality") +
  ggplot2::guides(alpha = "none") +
  ggplot2::scale_fill_manual(
    values = c("Bad" = "red", "Good" = "green", "Unknown" = "blue")
  ) +
  ggplot2::facet_grid(ImpExp ~ Reporter, scales = "free", space = "free_x")
```

Plot showing years covered by `expanded_trade_sources`, colored by scanned/not scanned:
```{r, eval = has_ggplot2, fig.alt="Plot showing years covered by expanded_trade_sources, colored by scanned/not scanned"}
ggplot2::ggplot(expanded_trade_sources, ggplot2::aes(y = Trade, x = Year)) +
  ggplot2::geom_tile(ggplot2::aes(fill = factor(Scanned), alpha = .8)) +
  ggplot2::theme_dark() +
  ggplot2::labs(title = "Scanned sources_trade by Country", fill = "Scanned") +
  ggplot2::scale_fill_manual(
    labels = c("0" = "No", "1" = "Yes"), values = c("0" = "red", "1" = "green")
  ) +
  ggplot2::guides(alpha = "none") +
  ggplot2::facet_grid(ImpExp ~ Reporter, scales = "free", space = "free_x")
```

Plot showing years covered by `expanded_trade_sources`, colored by whether or not in SACO:
```{r, eval = has_ggplot2, fig.alt="Plot showing years covered by expanded_trade_sources, colored by whether or not in SACO"}
ggplot2::ggplot(
  expanded_trade_sources,
  ggplot2::aes(y = Trade, x = Year, alpha = .8)
) +
  ggplot2::geom_tile(ggplot2::aes(fill = factor(In_Saco))) +
  ggplot2::theme_dark() +
  ggplot2::labs(title = "Source Availability by Country", fill = "In Saco") +
  ggplot2::guides(alpha = "none") +
  ggplot2::scale_fill_manual(
    labels = c("0" = "No", "1" = "Yes"), values = c("0" = "red", "1" = "green")
  ) +
  ggplot2::facet_grid(ImpExp ~ Reporter, scales = "free", space = "free_x")
```
